How Slow do my Transient Typechecks Go?

### 2019-08-17 graphs/preamble.tex
Reworded half the paper

\pgfplotsset{%
% WARNING: do not put empty lines here, and make sure each = ... is terminated with a comma!
	every axis/.append style = {% options for all axes/graphs
		cycle list={}, tick align=center, xtick pos = bottom, enlarge x limits=false, ymin=0, axis on top,%
		every axis plot/.style={GraphNormal},%
		every tick label/.append style={font=\tiny},%
		every axis label/.append style={font=\tiny},%
	},%
	%
	Full Left/.style = {%
		ymax={#1}, axis y line*=left, height=3.2cm, width=5.9cm, try min ticks=5},%
	Right Axis/.style = {%
		axis y line*=right, axis x line=none, xmin=0, xmax=100},%
	%
	Scatter/.style = {%
		every axis plot/.append style={only marks}, mark=o,
		xticklabel={$\pgfmathprintnumber{\tick}$\%}},%
	%
	Full Scatter Left/.style = {Scatter, Full Left={#1},
		enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
	%
	Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
	% Variants of the above that contain axis labels
	Full Scatter Left Edge/.style   = {Full Scatter Left={#1},        ylabel={Average Time (ms)}},%
	Full Scatter Left Bottom/.style = {Full Scatter Left={#1},        xlabel={Type Annotations}},%
	Full Scatter Left Corner/.style = {Full Scatter Left Bottom={#1}, Full Scatter Left Edge={#1}},%
	Full Scatter Right Edge/.style  = {Full Scatter Right={#1},       ylabel={Relative to 0\% Typed}},%
	%
	%
	Column/.style n args = {2}{%
		const plot, every axis plot/.append style={draw=GraphNormal, no markers, fill=GraphNormal!40}},%
	%
	Full Column Left/.style 2 args        = {Column, Full Left={#1},           xtick distance={#2}},%
	Full Column Right/.style              = {Full Column Left={#1}{1},         Right Axis},%
	% Variants of the above that contain axis labels
	Full Column Left Edge/.style 2 args   = {Full Column Left={#1}{#2},        ylabel={Average Time (ms)}},%
	Full Column Left Bottom/.style 2 args = {Full Column Left={#1}{#2},        xlabel={Type Annotation Index}},%
	Full Column Left Corner/.style 2 args = {Full Column Left Bottom={#1}{#2}, Full Column Left Edge={#1}{#2}},%
	Full Column Right Edge/.style         = {Full Column Right={#1},           ylabel={Relative to Untyped}},%
}

>>>>

\colorlet{GraphBase}{gray} % The colours for no type annotation
\colorlet{GraphA}{ACMRed}
\colorlet{GraphB}{ACMBlue}
\colorlet{GraphAB}{ACMPurple}

\pgfplotsset{%
% WARNING: do not put empty lines here, and make sure each = ... is terminated with a comma!
	every axis/.append style = {% options for all axes/graphs
		cycle list={}, tick align=center, xtick pos = bottom, ymin=0, axis on top,%
		axis y line*=left, try min ticks=5,%
		every tick label/.append style={font=\tiny},%
		every axis label/.append style={font=\tiny},%
	},%
	%
	Full/.style = { % For the page of graphs for each benchmark
		enlarge x limits={abs=2pt}, ymax={#1}, height=3.2cm, width=5.9cm,%
		every axis plot/.style={mark size=1pt, draw=GraphNormal, fill=GraphNormal!40}},%
	Right Axis/.style = { % For right axes
		axis y line*=right, axis x line=none, xmin=0, xmax=100},% 
	%
	Scatter/.style = { % For the scatter plots
		xticklabel={$\pgfmathprintnumber{\tick}$\%},%
		every axis plot/.append style={only marks}, mark=o},%
	%
	% Variants of the above that contain axis labels
	Scatter Left Label/.style   = {ylabel={Average Time (ms)}},%
	Scatter Bottom Label/.style = {xlabel={Type Annotations}},%
	Scatter Right Label/.style  = {ylabel={Relative to 0\% Typed}},%
	%
	%
	Column/.style = {% For the column graphs
		enlarge x limits=false, xtick distance={#1}, const plot,%
		every axis plot/.append style={no markers}},%
	%
	% Variants of the above that contain axis labels
	Column Left Label/.style   = {ylabel={Average Time (ms)}},%
	Column Bottom Label/.style = {xlabel={Type Annotation Index}},%
	Column Right Label/.style  = {ylabel={Relative to Untyped}},%
	%
	Pattern/.style = { % For the page of graphs for each benchmark
		enlarge x limits={abs=1.5pt}, ymax={#1}, height=3.2cm, width=4.4cm, %
		every axis plot/.style={mark size=0.75pt, draw=GraphBase, fill=GraphBase!40}},%
	%
	Pattern A/.style = {draw=GraphA, fill=GraphA!40, mark=*}, %
	Pattern B/.style = {draw=GraphB, fill=GraphB!40, mark=*}, %
	Pattern AB/.style = {draw=GraphAB, fill=GraphAB!40, mark=*}, %
	%Pattern Left/.style = {Left Axis={#1}, height=3.2cm, width=7.9cm}, % Size for pattern graphs
	%Full Scatter Left/.style = {Scatter, Full Left={#1},
	%	enlarge x limits={abs=2pt}, every axis plot/.append style={mark size={1pt}}},%
	%Full Scatter Right/.style       = {Full Scatter Left={#1},        Right Axis},%
	%sleft = "Pattern Scatter Left"
	%cleft = "Pattern Column Left"
	%sright = "Pattern Scatter Right"
	%cright = "Pattern Column Right"
	%Pattern Scatter A, Pattern Scatter B
}

### 2019-08-17 paper.tex
Reworded half the paper

\begin{abstract}
Just-in-time compilation and optimisation in virtual machines can
eliminate much of the overhead of transient typechecks.  Unfortunately
the improvement is not uniform: while most typechecks can be optimised
away, some typechecks will significantly decrease a program's
performance.  In this paper we investigate how benchmark results
indicate which typechecks cause the most problems.
Using these techniques,
programmers can optimise programs by removing typechecks,
and VM engineers can identify opportunities for compiler optimisation.

@@

%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
%%KJX XML IS BROKEN!!!
\begin{CCSXML}
\end{CCSXML}

@@

Gradual Typing aims to add static type annotations to dynamic languages, increasing
their safety while maintaining flexibility
\citep{GiladPluggable2004,Siek2006,XXXSiek2015}
--- or, complementarily,
to permit dynamic type annotations within static languages, increasing
flexibility while maintaining safety
\citep{AbadiTOPLAS1991}.

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end\,---\,``pluggable types'' as in Strongtalk \cite{strongtalk} or ``erasure
semantics'' as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript
\citep{typeScriptECOOP}\,---\,
%
all types are erased before the execution, limiting the benefit of
types to the statically typed parts of programs, and preventing
programs from depending on type checks at run time.  In the middle,
``transient'' or ``type-tag'' checks as in Reticulated Python 
% and ``concrete'' checks as in Thorn
offer first-order semantics, checking
whether an object's type constructor or supported methods match
explicit type declarations
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports an alternative ``monotonic'' semantics
which mutates an object to narrow its concrete type when it is passed
into a more specific type context.
At the other end of the spectrum, behavioral
typechecks as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14},
and Reticulated Python's proxies,
support higher-order semantics, retaining
types until run 
time, performing the checks eagerly, and giving detailed information
about type violations as soon as possible via blame
tracking \cite{blame2009,blameForAll2011}.
Finally, Ductile typing
dynamically interprets a static type system at run time
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
Unfortunately, any gradual system with run-time semantics
(i.e.\ everything more complex than erasure) currently
imposes a significant run-time performance overhead to provide those
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other hand developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

This paper builds upon our recent work on optimising transient
typechecks \cite{Roberts2017,roberts-and-co-ecoop-2019} to
investigate how benchmark results can be repurposed to identify which
particular transient typechecks are resistant to optimisation.
The next section discusses dynamic typechecks and gradual typing in
Grace, then section~\ref{s-eval} describes the benchmarking
protocols. Section~\ref{s-overall} then presents the results of
analysing overall benchmark results, while section~\ref{s-individual}
looks at the results of benchmarking individual type checks.
Finally~\ref{s-concl} discusses our results, briefly considers threats
to validity, and concludes.

@@

Our work is based on the Moth virtual machine 
\cite{Roberts2017,roberts-and-co-ecoop-2019},
an implementation
of the Grace programming language 
\citep{graceOnward12,graceSigcse13},
based on the Graal and Truffle toolchain
\cite{Wurthinger:2017:PPE,Wurthinger2013}
and developed from a Newspeak implementation based on the  Simple
Object Machine \cite{Daloze2016,SOMns}.

\subsection{Grace's Transient Type Checking}

Grace is an object-oriented, imperative, educational programming
language, with a focus on introductory programming
courses, but also intended for more advanced study and research \citep{graceOnward12,graceSigcse13}.
%
While Grace's syntax draws
from the so-called ``curly bracket'' tradition of C, Java, and
JavaScript, the structure of the language
is in many ways closer to Smalltalk:
all computation is via dynamically dispatched  ``method requests''
where the object receiving the request decides which code to run,
and
%
returns within lambdas are ``non-local'', returning to the method
activation in which the block is instantiated \citep{bluebook}.  In
other ways, Grace is closer to JavaScript than Smalltalk: Grace
objects are created from object literals, rather than by
instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
objects and classes can be deeply nested within each 
other \citep{betabook}.

\paragraph{Grace's Typing}
Critically, Grace's declarations and methods' arguments
and results can be annotated with types, and those types can be  checked
either statically or dynamically. This means the type system is
intrinsically gradual:%
%
~type annotations should not affect the semantics of a correct
program \citep{XXXSiek2015}, and the type system
includes a distinguished ``{Unknown}'' type which matches any other type
and is the implicit type for untyped parts of the program.

The static core of Grace's type system is well-described
elsewhere \citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object implements a type whenever it
implements all the methods required by that type,
rather than requiring classes or objects to declare types explicitly \RR{(explicitly or nominally)?}. 
% This should say Moth's, not Grace's, but I know there is objection to that
Grace's transient dynamic typechecks are first-order, so methods match when they have
the same name and arity, while
argument and return types are ignored.
A type thus expresses the requests an object can respond to,
for example whether a particular accessor is available,
rather than a nominal location in an explicit inheritance hierarchy.

% Ditto
Grace then checks the types of values at run time:
%
\begin{itemize}
\item the values of arguments are checked after a method is requested, 
      but before the body of the method is executed;
\item the value returned by a method is checked after its body is executed; and
\item the values of variables are checked
      whenever written or read by user code.
\end{itemize}

To see how this works in practice, consider this code snippet, which
will run in Grace.

@@

Dynamic typechecks will take place:

@@

\item on line 7,
when object \code{o} initialises the variable \code{t},
we check it has the method \code{three};

\item on line 8,
when the value of \code{t} is read,
it must again have the method \code{three};

\item on line 2,
when the \code{three} method requested by ``\code{t.three}'' returns,
the return value must be a \code{Number};
and (presumably) within the definition of
%
\code{printNumber(n :   Number)}
%
(not shown) to check that its argument is a number. 
\end{itemize}

Note that we never check
whether e.g.\ the result of requesting ``\code{t.three}'' is actually
a \code{String} (as one may expect from line 5) because we check
only first-order types (whether objects have conforming methods) not higher-order
types (whether those methods' argument and result types conform), and
we check only when values flow through explicit type annotations.
This is why the type declared in lines 4-6 is checked only on line 7
(where it is mentioned explicitly); and the check only requires the
presence of a method called \code{three}, regardless of the method's
declared return type.

@@

\section{Related Work}

>>>>

\begin{abstract}
One form of type checking used in gradually typed language is \emph{transient type checking}: when ever an object 'flows' through code with a type annotation, the object is dynamically checked to have the methods required by said annotation.

Though na\"ive implementations of this have a high runtime overhead, just-in-time compilation and optimisation in virtual machines can eliminate much of this overhead.  Unfortunately
the improvement is not uniform: while most type checks can be optimised
away, some will significantly decrease a program's
performance, and some may even increase it.  In this paper we investigate how benchmark results can be used to identify which type checks have the greatest effects on performance.
Using these techniques,
programmers can optimise programs by removing expensive type checks,
and VM engineers can identify new opportunities for compiler optimisation.

@@

%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011041.10011044</concept_id>
<concept_desc>Software and its engineering~Just-in-time compilers</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10011011</concept_id>
<concept_desc>Software and its engineering~Object oriented languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011041.10010943</concept_id>
<concept_desc>Software and its engineering~Interpreters</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

@@

Gradual Typing aims to add static type annotations to dynamic languages, increasing
their safety while maintaining flexibility
\citep{GiladPluggable2004,Siek2006,XXXSiek2015}
--- or, complementarily,
to permit dynamic type annotations within static languages, increasing
flexibility whilst maintaining safety
\citep{AbadiTOPLAS1991}.

There is a spectrum of different approaches to gradual typing
\cite{kafka18,bensurvey18icfp}.
At one end is ``pluggable types'' (as in Strongtalk \cite{strongtalk}) or ``erasure
semantics'' (as in 
%% Two important approaches are optional\citep{}
%% and gradual typing\citep{GiladPluggable2004,Siek2006,XXXSiek2015}.
%% These are applied to dynamic languages to reap the benefits of typing, but
%% unfortunately also have limitations.
%With optional or pluggable approaches such as 
%\citeurl{TypeScript,}{TypeScript}{Microsoft}{27 June
%2018}{https://www.typescriptlang.org/}
TypeScript
\citep{typeScriptECOOP}), here
%
all types are erased before execution, limiting the benefit of
types to statically typed portions of programs, whilst not providing 
any for of run time checking.  In the middle there are
``transient'' or ``type-tag'' checks (as in Reticulated Python)
% and ``concrete'' checks as in Thorn
, these offer first-order semantics, by checking
that an object's type constructor or supported methods match
any explicit type annotations that the object flows through.
%%%WRONG kjx
%, and checking only when data flows through those declarations
\cite{Siek2007,Bloom2009,concrete15,reticPython2014,Greenman2018}.
Reticulated Python also supports alternative ``monotonic'' semantics
which mutate objects by narrowing their concrete types when passed
into contexts with more specific types.
On the other end of the spectrum are behavioural
typechecks (as in Typed Racket \cite{typedScheme08,takikawa2012},
Gradualtalk \cite{gradualtalk14}),
and ``proxies'' (provided by Reticulated Python), these
support higher-order semantics, retaining
types until run 
time, and performing checks eagerly, thus detailed information
about type violations can be given as soon as possible via ``blame''
tracking \cite{blame2009,blameForAll2011}.
Finally, there is also ductile typing,
which dynamically interprets a static type system at run time
\cite{Ductile2011}.
%
%\mwh{Does a monotonic semantics fit on this spectrum somewhere?}
%\kjx{Not in the greenman/fellenstein spectrum, but I guess I'd put it
%  after concrete and before behavioural. Does it matter?}
%
% This said "any", but the ecoop paper specifically showed that 
Unfortunately, most gradual system with run-time semantics
(as opposed to type erasure as in TypeScript) currently
imposes a significant run-time performance overhead to provide such
semantics.  This has lead to a significant body of research,
on one hand developing techniques to optimise gradual typing 
\citep{Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Greenman2018}
and on the other, developing techniques to evaluate the
resulting performance  \cite{Takikawa2016,Greenman2019jfp}.

This paper builds upon our recent work on optimising transient
type checks \cite{Roberts2017,roberts-and-co-ecoop-2019} by
investigating how benchmark results can be repurposed to identify which
particular transient type checks are resistant to optimisation.
The next section discusses dynamic type checks and gradual typing in
Moth (an implementation of the Grace language), then section~\ref{s-eval} describes our benchmarking protocol. Section~\ref{s-overall} then presents the overall results of
our benchmarks, while section~\ref{s-individual}
looks at the results of benchmarking individual type checks.
Section~\ref{s-rel} presents some additional related work, and finally \ref{s-concl} summarises our results, and briefly considers threats to validity.

@@

Our work is based on the Moth virtual machine 
\cite{Roberts2017,roberts-and-co-ecoop-2019},
an implementation
of the Grace programming language 
\citep{graceOnward12,graceSigcse13}.
Moth is based on the Graal and Truffle toolchain
\cite{Wurthinger:2017:PPE,Wurthinger2013},
and developed from a Newspeak implementation based on the  Simple
Object Machine \cite{Daloze2016,SOMns}.

\subsection{Grace and Transient Type Checking}

Grace is an object-oriented, imperative, educational programming
language, with a focus on introductory programming
courses, but also intended for more advanced study and research \citep{graceOnward12,graceSigcse13}.
%
While Grace's syntax draws
from the so-called ``curly bracket'' traditions of C, Java, and
JavaScript, the structure of the language
is in many ways closer to Smalltalk:
all computation is done via dynamically dispatched  ``method requests''
where the object receiving the request decides which they return to the point where execution first encountered the lambda \citep{bluebook}.  In
other ways, Grace is closer to JavaScript than Smalltalk: Grace
objects are created from object literals, rather than by
instantiating classes \citep{Black2007-emeraldHOPL,JonesECOOP2016} and
objects and classes can be deeply nested within each 
other \citep{betabook}.

\paragraph{Grace's Typing}
Critically, in Grace, declarations, methods can be annotated with types.
As Grace is designed to support a variety of teaching methods, implementation of Grace are free to check such type annotations  statically, dynamically, or not at all.
 Thus the type system of Grace is intrinsically gradual:%
%
~type annotations should not affect the semantics of a correct
program \citep{XXXSiek2015}, and the type system
includes a distinguished ``{Unknown}'' type which matches any other type,
and is the default type when type annotations are omitted.

Static typing for the core of Grace's type system has been described
elsewhere \citep{TimJonesThesis};
here we explain how
these types can be understood 
dynamically, from the Grace programmer's point of view.
Grace's types are structural \citep{graceOnward12},
that is, an object conforms to a type whenever it conforms to the ''structural`` requirements of a type,
rather than requiring classes or objects to explicitly declare their intended type.

In Grace, types specify a set of method signatures that an object must provide. Thus a type expresses the requests an object can respond to, for example whether a particular accessor is available, rather than a location in a class hierarchy.

Moth's implementation of transient type checks however are only only first-order, as such it only dynamically checks that an object has methods of the same name and arity as are required by a type, however any argument and return types of such methods are not checked.

In particular, Moth performs such type checks are run time:
\begin{itemize}
\item when a method is requested, arguments that are passed are checked against the corresponding parameter type annotations of the called method, this is done before the body of the method is executed;
\item when the body of a method has finished executing, but before it returns to its caller, the method's return value is checked against the return type annotation of the called method;
\item whenever a variable is read or written to, its value is checked against the type specified by the variables declaration.
\end{itemize}

To see how this works in practice, consider this piece of Grace code:

@@

Moth will perform dynamic type checks:

@@

\item on line 7,
when the \code{o} object initialises the variable \code{t},
Moth checks that \code{o} has a 0-argument method called \code{three};

\item on line 8,
when the value of \code{t} is read,
Moth checks that it's value (\code{o}) still has a \code{three} method;

\item on line 2,
when the \code{three} method requested by ``\code{t.three}'' returns,
Moth checks that returned value conforms to the \code{Number} type;
and (presumably) within the definition of
%
\code{printNumber(n :   Number)}
%
(not shown), Moth will again check that the value is a \code{Number}.
\end{itemize}

Note that we never check
whether the result of requesting ``\code{t.three}'' is actually
a \code{String} (as one may expect from line 5) because Moth only performs first-order type checks
(it checks whether objects have conforming methods) not higher-order
checks (whether the argument and result types of methods' conform). In addition, Moth
only checks when values flow through explicit type annotations.
This is why the type declared in lines 4-6 is checked only on line 7
(where it is mentioned explicitly); and the check only requires the
presence of a method called \code{three}, regardless of the method's
declared return type.

@@

\section{Related Work}
\label{s-rel}
